### 2020-06-26 submission/manuscript.Rmd
Additional edits and incorporate Josh's suggestion

## Abstract
The microbiota plays a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs). However, much of the mechanistic work examining CDIs in mouse models use a single university colony or vendor. We treated mice from 6 different colony sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by *C. difficile* challenge 1 day later and measured *C. difficile* colonization levels through 9 days post-infection. The microbiota was profiled via 16S rRNA gene sequencing analysis to examine variation across colony sources and alterations due to clindamycin treatment and *C. difficile* challenge. While all sources of mice were colonized 1-day post-infection, variation in *C. difficile* colonization levels emerged from days 3-7 post-infection with 3 sources colonized with *C. difficile* for slightly longer and at higher levels. We identified bacterial taxa with different relative abundances across colony sources throughout the experiment, as well as taxa that were consistently impacted by clindamycin treatment in all sources of mice. We created logistic regression models that successfully classified mice based on whether they cleared *C. difficile* by 7 days post-infection using baseline, post-clindamycin, and post-infection community composition data. After examining the taxa that were most important to the classification models, we identified a subset of key taxa that varied across colony sources (*Bacteroides*, *Deferribacteraceae*), were altered by clindamycin (*Porphyromonadaceae*, *Ruminococcaceae*), or both (*Enterobacteriaceae*, *Enterococcus*, *Bifidobacteriaceae*, *Coriobacteriaceae*, *Lachnospiraceae*, and *Verrucomicrobiaceae*). These results suggest the response of the initial gut microbiota to clindamycin treatment influences *C. difficile* 630 colonization dynamics.

@@

In the past, our group has attempted to introduce more microbiome variation into the CDI mouse model by using a variety of antibiotic treatments [@Reeves2011; @Schubert2015; @Jenior2017; @Jenior2018]. An alternative approach to maximize microbiome variation is to use mice from multiple sources [@Velazquez2019; @Osbelt2020]. Microbiome differences between different mouse vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018]. Additionally, different research groups have observed different CDI outcomes in mice despite using similar models and the microbiome has been proposed as one factor potentially mediating susceptibility [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014]. Here we examined how variations in the baseline microbiome and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C. difficile* colonization and the time needed to clear the infection.

@@

**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization regardless of colony source.** To test how the microbiotas of mice from different colony sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan (the Young and Schloss lab colonies), the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019]. After a 13-day acclimation period for the mice ordered from vendors, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 1A). Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared the CDI within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. The day after infection, *C. difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of colony source (Fig. 1B). Interestingly, variation in *C. difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 1B and Table S1), suggesting mouse colony source is associated with *C. difficile* clearance. We conducted two experiments approximately 3 months apart and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig. S1A-B, Fig. 1D). Although *C. difficile* 630 causes mild symptoms in mice comparied to other *C. difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig. 1C, E and Table S2). Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig. 1B-E), when `r percent_colonized_d7` of the mice were still colonized with *C. difficile* (Fig. S1C). By 9 days post-infection the majority of the mice from all sources had cleared *C. difficile* (Fig. S1C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment. Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment. Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of colony source, but variation across sources emerged with 3 out of 6 sources taking longer to clear *C. difficile*.

@@

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```
**Bacterial communities consistently vary across mouse colony sources despite antibiotic and infection perturbations.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C. difficile* clearance could be explained by microbiota variation across the 6 sources. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment. Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice. Examining the bacterial communities at baseline, prior to clindamycin treatment there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig. 2A-B and Table S3). As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and communities started to recover 1 day post-infection (Fig. 2C-D). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C. difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other groups (Fig 2C-F and Table S4). While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C. difficile* faster, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across vendors.
```{r adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

@@

source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
```
Next, we compared the bacterial communities from the 6 colonies over the course of the experiment using principal coordinate analysis (PCoA) of the Theta YC distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed colony source was the major factor explaining the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment (Movie S1 and Table S5). Since, the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the bacterial communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1). For all 3 timepoints, source and the interaction with cage significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 3 and Table S6). We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2 and Table S7), although most of the vendors also clustered by experiment, suggesting there was some community variation between the 2 experiments within each vendor. Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities. Importantly with the exception of the 2 University colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C. difficile* challenge.

Since there was some variation in microbiota communities between experiments at baseline, we next looked at how similar the communities were within the same source and between sources in response to clindamycin treatment and *C. difficile* challenge (Fig. 4) . The baseline communities varied most between experiments for Schloss, Young, and Envigo mice and variation between sources of mice was high (Fig. 4A). Clindamycin treatment reduced variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished (Fig. 4B). Post-infection, the community variation started to increase within sources of mice and variation between sources of mice started to return (Fig. 4C). By using mice from multiple sources we were able to increase the number of microbiota communities we tested with the clindamycin *C. difficile* colonization mouse model.
```{r source_taxa}
#Number of different otus and families across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_families <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
source_families_dn1 <- source_families %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_families_d0 <- source_families %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
source_families_d1 <- source_families %>% filter(day == 1) %>% nrow()

@@

#Families shared across baseline, post-clindamycin, and post-infection timepoints
source_f_dn1 <- source_families %>% filter(day == -1) %>% pull(family)
source_f_d0 <- source_families %>% filter(day == 0) %>% pull(family)
source_f_d1 <- source_families %>% filter(day == 1) %>% pull(family)
shared_source_families_dn1to1 <- length(intersect_all(source_f_dn1, source_f_d0, source_f_d1))
```
After finding differences at the community level, we next identified the taxa that varied across sources of mice over the initial days of the experiment. We examined bacterial relative abundances at the operational taxonomic unit (OTU) and family levels, expecting the number of differences to be reduced at the family level due to the nature of bacterial taxonomy [@Nguyen2015]. Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and `r source_families_dn1` families (Table S8-9) with relative abundances that varied across colony sources. Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Table S8-9). After *C. difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Table S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment. In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig. 5D-F). Importantly, some of the taxa that consistently varied across sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment. In summary, mouse bacterial communities significantly varied according to colony source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C. difficile* challenge.

@@

clindamycin_families <- read_tsv("../data/process/family_stats_dn1to0.tsv")
clindamycin_families_sig <- clindamycin_families %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters a subset of taxa that were found in all colony sources.** Although there were bacteria that consistently varied across colony sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of colony source. By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs and `r clindamycin_families_sig` families that were altered after clindamycin treatment (Fig. 6 and Table S10-11). Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) and 3 families (*Porphyromonadaceae*, *Enterococcaceae*, *Lachnospiraceae*) overlapped (Fig. 5, Fig. 6C-D). These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all colony sources and only a subset of the taxa also varied across colony sources.

@@

#Venn diagram calculations for OTUs----
venn_otus_source <- length(dayn1to1_and_interp_combined)
venn_otus_intersect <- length(intersect(dayn1to1_and_interp_combined, paired_and_interp_combined))
venn_otus_clind <-  length(paired_and_interp_combined)
#Venn diagram calculations for families
venn_families_source <- length(dayn1to1_and_interp_combined_f)
venn_families_intersect <- length(intersect(dayn1to1_and_interp_combined_f, paired_and_interp_combined_f))
venn_families_clind <-  length(paired_and_interp_combined_f)
```
**Source-specific and clindamycin impacted bacteria distinguish *C. difficile* colonization status in mice.** After identifying taxa that varied by colony source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C. difficile* colonization at day 7 (Fig. 1D, Fig. S1C). We trained L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C. difficile* colonization status on day 7 (Fig. S3A-B). All models were better at predicting *C. difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12), however the models trained with OTU level data generally performed better than those trained with family level data with the expception of the models based on the post-infection (day 1) communities (Fig. S3C-D). Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an AUROC of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13). Thus, we were able to use community bacterial relative abundance data alone to differentiate mice that had cleared *C. difficile* before day 7 from the mice still colonized with *C. difficile*. Interestingly, the model built with OTU relative abundance data post-clindamycin treatment had the best performance, suggesting how the bacterial community responds to clindamycin treatment has the greatest influence on subsequent *C. difficile* colonization dynamics.

Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 taxa that had the highest absolute feature weights in each of the 6 models (Table S14-15). First, we looked at OTUs from the model with the best performance that was based on the post-clindamycin treatment bacterial community data. While most of the 20 OTUs had low relative abundances on day 0, *Enterobacteriace*, *Bacteroides* and *Proteus* had high relative abundances in at least one source of mice and significantly varied across sources (Fig. 7A). Next, the top 20 taxa from each model were compared to the list of taxa that varied across colony source (Fig. 5 and Table S8-9) at the same timepoint and the taxa that were altered by clindamycin treatment (Fig. 6 and Table S10-11). We found a subset of OTUs and families that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig. S4, S5A-C). Combining the overall results for the 3 OTU models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig. 7B). Combining the overall results for the 3 family models identified `r venn_families_source` families associated with source, `r venn_families_clind` families associated with clindamycin treatment and `r venn_families_intersect` families associated with both (Fig. S5D). Several OTUs (*`r list_key_otus`*) and families (*`r list_key_families`*) appeared across at least 2 models, so we examined how the relative abundances of these key taxa varied over the course of the experiment (Fig. 8 and Fig. S6). Throughout the experiment, there was at least 1 timepoint where relative abundances of these taxa significantly varied across sources (Table S16-17). Interestingly, there were no taxa that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C. difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection. Together, these results suggest the initial bacterial communities and their responses to clindamycin have a large influence on the time needed to clear *C. difficile*.

>>>>

## Abstract
The microbiota plays a key role in determining susceptibility to *Clostridioides difficile* infections (CDIs). However, much of the mechanistic work examining CDIs in mouse models use a single university colony or vendor. We treated mice from 6 different colony sources (2 University of Michigan colonies and 4 vendors) with a single clindamycin dose, followed by *C. difficile* challenge 1 day later and measured *C. difficile* colonization levels through 9 days post-infection. The microbiota was profiled via 16S rRNA gene sequencing analysis to examine variation across colony sources and alterations due to clindamycin treatment and *C. difficile* challenge. While all sources of mice were colonized 1-day post-infection, variation in *C. difficile* colonization levels emerged from days 3-7 post-infection with some sources colonized with *C. difficile* for slightly longer and at higher levels. We identified bacterial taxa with different relative abundances across colony sources throughout the experiment, as well as taxa that were consistently impacted by clindamycin treatment in all sources of mice. We created logistic regression models that successfully classified mice based on whether they cleared *C. difficile* by 7 days post-infection using baseline, post-clindamycin, and post-infection community composition data. After examining the taxa that were most important to the classification models, we identified a subset of key taxa that varied across colony sources (*Bacteroides*, *Deferribacteraceae*), were altered by clindamycin (*Porphyromonadaceae*, *Ruminococcaceae*), or both (*Enterobacteriaceae*, *Enterococcus*, *Bifidobacteriaceae*, *Coriobacteriaceae*, *Lachnospiraceae*, and *Verrucomicrobiaceae*). These results suggest the response of the initial gut microbiota to clindamycin treatment influences *C. difficile* 630 colonization dynamics.

@@

In the past, our group has attempted to introduce more microbiome variation into the CDI mouse model by using a variety of antibiotic treatments [@Reeves2011; @Schubert2015; @Jenior2017; @Jenior2018]. An alternative approach to maximize microbiome variation is to use mice from multiple sources [@Velazquez2019; @Osbelt2020]. Microbiome differences between different mouse vendors have been well documented and shown to influence susceptibility to a variety of diseases [@Stough2016; @Alegre2019], including enteric infections [@EtienneMesmin2017; @Velazquez2019; @Osbelt2020; @Lai2020; @Thiemann2017; @Rolig2013; @Ge2018]. Additionally, different research groups have observed different CDI outcomes in mice despite using similar models and the microbiome has been proposed as one factor potentially mediating CDI susceptibility and outcomes [@Hutton2014; @Lawley2013; @Reeves2011; @Buffie2011; @Jenior2018; @Buffie2014]. Here we examined how variations in the baseline microbiome and responses to clindamycin treatment in C57BL/6 mice from six different sources influenced susceptibility to *C. difficile* colonization and the time needed to clear the infection.

@@

**Clindamycin treatment renders all mice susceptible to *C. difficile* 630 colonization regardless of colony source.** To test how the microbiotas of mice from different colony sources impact colonization dynamics after clindamycin exposure, we utilized C57BL/6 mice from 6 different sources: two colonies from the University of Michigan (the Young and Schloss lab colonies), the Jackson Laboratory, Charles River Laboratories, Taconic Biosciences, and Envigo (which was formerly Harlan). These 4 vendors were chosen because they represent commonly used vendors for CDI studies in mice [@Spinler2016; @Markey2018; @McKee2018; @Yamaguchi2020; @Stroke2018; @EtienneMesmin2017; @Quigley2019; @Mullish2019]. After a 13-day acclimation period for the mice ordered from vendors, all mice were treated with 10 mg/kg clindamycin via intraperitoneal injection and one day later challenged with 10^3^ *C. difficile* 630 spores (Fig. 1A). Clindamycin was chosen because we have previously demonstrated mice are rendered susceptible, but consistently cleared the CDI within 9 days [@Jenior2018; @Tomkovich2019], clindamycin is frequently implicated with human CDIs [@Guh2018], and is also part of the antibiotic treatment for the frequently cited 2008 CDI mouse model [@Chen2008]. The day after infection, *C. difficile* was detectable in all mice at a similar level (median CFU range: `r format(min_median_d1, digits = 2, scientific=TRUE)`-`r format(max_median_d1, scientific=TRUE, digits = 2)`; *P*~FDR~ = `r padj_d1_cfu`), indicating clindamycin rendered all mice susceptible regardless of colony source (Fig. 1B). Interestingly, variation in *C. difficile* CFU levels across sources of mice emerged from days 3-7 post-infection (all *P*~FDR~ $\le$ `r max_padj_d3to7`; Fig. 1B and Table S1), suggesting mouse colony source is associated with *C. difficile* clearance. We conducted two experiments approximately 3 months apart and while the colonization dynamics were similar across most sources of mice, there was some variation between the 2 experiments, particularly for the Schloss and Envigo mice (Fig. S1A-B). Although *C. difficile* 630 causes mild symptoms in mice comparied to other *C. difficile* strains [@Theriot2011], we also saw that weight change significantly varied across sources of mice with the most weight lost two days post-infection (Fig. 1C and Table S2). Importantly, there was also one Jackson and one Envigo mouse that died between 1- and 3-days post-infection during the second experiment. Interestingly, mice ordered from Jackson, Taconic, and Envigo tended to lose more weight (although there was variation between experiments with Schloss and Envigo mice), have higher *C. difficile* CFU levels and take longer to clear the infection compared to the other sources of mice, which was particularly evident 7 days post-infection (Fig. 1B-C, Fig. S2C-D), when `r format(percent_colonized_d7, digits = 1)`% of the mice were still colonized with *C. difficile* (Fig. S1E). By 9 days post-infection the majority of the mice from all sources had cleared *C. difficile* (Fig. S1C) with the exception of 1 Taconic mouse from the first experiment and 2 Envigo mice from the second experiment. Thus, clindamycin rendered all mice susceptible to *C. difficile* 630 colonization, regardless of colony source, but variation across sources emerged with 3 out of 6 sources taking longer to clear *C. difficile*.

@@

shannon_stats_all <- read_tsv("../data/process/shannon_stats_all_days.tsv")
padj_dn1_shannon <- shannon_stats_all %>% filter(day == -1) %>% pull(p.value.adj)
```
**Bacterial communities consistently vary across mouse colony sources despite antibiotic and infection perturbations.** Given the well known variation in mouse microbiomes across vendors and university colonies [@Alegre2019], we hypothesized that the variation in *C. difficile* clearance could be explained by microbiota variation across the 6 sources. We used 16S rRNA gene sequencing to characterize the fecal bacterial communities from the mice over the course of the experiment. Since antibiotics and other risk factors of CDIs are associated with decreased microbiota diversity [@Ross2016], we first examined alpha diversity measures across the 6 sources of mice. Examining the bacterial communities at baseline, prior to clindamycin treatment there was a significant difference in the number of observed OTUs (*P*~FDR~ = `r padj_dn1_richness`), but not Shannon diversity index (*P*~FDR~ = `r padj_dn1_shannon`) across sources of mice (Fig. 2A-B and Table S3). As expected, clindamycin treatment decreased richness and Shannon diversity across all sources of mice, and communities started to recover 1 day post-infection (Fig. 2C-D). Interestingly, significant differences in diversity metrics across sources (*P*~FDR~ < 0.05) emerged after both clindamycin and *C. difficile* infection, with Charles River mice having higher richness and Shannon Diversity than most of the other sources (Fig 2C-F and Table S4). While Charles River mice had more diverse microbiotas, Young and Schloss lab mice were also able to clear *C. difficile* faster, suggesting microbiota diversity alone does not explain the observed variation in *C. difficile* colonization across vendors.
```{r adonis}
adonis_stats_all <- read_tsv("../data/process/adonis_all.tsv")
r_sq_ordered_adonis <- adonis_stats_all %>% arrange(desc(r_sq))
rsq_adonis_source <- r_sq_ordered_adonis %>% pull(r_sq) %>% max
p_adonis_source <- r_sq_ordered_adonis %>% pull(p) %>% max

@@

source_adonis_dn1 <- read_tsv("../data/process/adonis_dn1_source.tsv")
adonis_schloss_rsq_sum <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(r_sq) %>% sum
p_adonis_schloss_max <- source_adonis_dn1 %>% filter(source == "Schloss") %>% select(p) %>% max
adonis_young_rsq_sum <- source_adonis_dn1 %>% filter(source == "Young") %>% select(r_sq) %>% sum
p_adonis_young_max <- source_adonis_dn1 %>% filter(source == "Young") %>% select(p) %>% max
```
Next, we compared the bacterial communities from the 6 colonies over the course of the experiment using principal coordinate analysis (PCoA) of the Theta YC distances. Permutational multivariate analysis of variance (PERMANOVA) analysis revealed colony source was the major factor explaining the observed variation across fecal communities (R^2^ = `r rsq_adonis_source`, *P* = `r p_adonis_source`) followed by interactions between cage and day of the experiment (Movie S1 and Table S5). Since, the majority of the perturbations happened over the initial days of the experiment, we decided to focus on the bacterial communities at baseline (day -1), after clindamycin treatment (day 0), and post-infection (day 1). For all 3 timepoints, source and the interaction with cage significantly explained most of the observed community variation (combined R^2^ = `r dn1_rsq_sum`, `r d0_rsq_sum`, `r d1_rsq_sum`, respectively; *P* = `r p_adonis_intial`; Fig. 3 and Table S6). We also compared baseline communities across the 2 experiments, and found experiment and cage significantly explained the observed variation only for the Schloss and Young lab mouse colonies (Fig. S2A-B and Table S7). However, most of the vendors also clustered by experiment (Fig. S2C-D, F), suggesting there was some community variation between the 2 experiments within each source. Thus, mouse colony source was the factor that explained the most variation observed in the bacterial communities. Importantly with the exception of the 2 University colonies, the community of each source clustered apart from one another suggesting each community had a unique response to clindamycin treatment and *C. difficile* challenge.

Since there was some variation in microbiota communities between experiments at baseline, we next looked at how similar the communities were within the same source and between sources in response to clindamycin treatment and *C. difficile* challenge (Fig. 4) . The baseline communities varied most between experiments for Schloss, Young, and Envigo mice and variation between sources of mice was high (Fig. 4A). Clindamycin treatment reduced variation between experiments within Schloss, Young and Jackson mice and some of the variation between sources diminished, particularly for the University of Michigan and Charles River mice (Fig. 4B). Post-infection, the community variation started to increase within sources of mice and variation between sources of mice started to increase towards previous levels (Fig. 4C). By using mice from multiple sources we were able to increase the number of microbiota communities we tested with the clindamycin *C. difficile* colonization mouse model.
```{r source_taxa}
#Number of different otus and families across sources at day -1, 0, and 1 timepoints
source_otus <- read_tsv("../data/process/otu_stats_dn1to1_combined.tsv")
source_otus_dn1 <- source_otus %>% filter(day == -1) %>% nrow()
source_families <- read_tsv("../data/process/family_stats_dn1to1_combined.tsv")
source_families_dn1 <- source_families %>% filter(day == -1) %>% nrow()
source_otus_d0 <- source_otus %>% filter(day == 0) %>% nrow()
source_families_d0 <- source_families %>% filter(day == 0) %>% nrow()
source_otus_d1 <- source_otus %>% filter(day == 1) %>% nrow()
source_families_d1 <- source_families %>% filter(day == 1) %>% nrow()

@@

#Families shared across baseline, post-clindamycin, and post-infection timepoints
source_f_dn1 <- source_families %>% filter(day == -1) %>% pull(family)
source_f_d0 <- source_families %>% filter(day == 0) %>% pull(family)
source_f_d1 <- source_families %>% filter(day == 1) %>% pull(family)
shared_source_families_dn1to1 <- length(intersect_all(source_f_dn1, source_f_d0, source_f_d1))
```
After finding differences at the community level, we next identified the taxa that varied across sources of mice over the initial days of the experiment. We examined bacterial relative abundances at the operational taxonomic unit (OTU) and family levels, expecting the number of differences to be reduced at the family level due to the nature of bacterial taxonomy [@Nguyen2015]. Focusing on the baseline communities first, there were `r source_otus_dn1` OTUs and `r source_families_dn1` families (Tables S8-9) with relative abundances that varied across colony sources. Clindamycin treatment reduced the number of taxa with relative abundances that varied across sources to `r source_otus_d0` OTUs and `r source_families_d0` families (Tables S8-9). After *C. difficile* challenge, there were `r source_otus_d1` OTUs and `r source_families_d1` families (Tables S8-9) with significantly different relative abundances across sources, as the communities started to recover from antibiotic treatment. In spite of the experimental perturbations that occurred during these 3 timepoints, there were `r shared_source_otus_dn1to1` OTUs (Fig 5A-C) and `r shared_source_families_dn1to1` families with relative abundances that consistently varied across colony sources (Fig. 5D-F). Importantly, some of the taxa that consistently varied across sources also shifted with clindamycin treatment. For example, *Proteus* increased after clindamycin treatment, but only in Taconic mice. *Enterococcus* was primarily found only in mice purchased from commercial vendors and also increased after clindamycin treatment. In summary, mouse bacterial communities significantly varied according to colony source throughout the course of the experiment and a consistent subset of bacterial taxa remained different across sources regardless of clindamycin and *C. difficile* challenge.

@@

clindamycin_families <- read_tsv("../data/process/family_stats_dn1to0.tsv")
clindamycin_families_sig <- clindamycin_families %>% filter(p.value.adj < 0.05) %>% nrow()
```
**Clindamycin treatment alters a subset of taxa that were found in all colony sources.** Although there were bacteria that consistently varied across colony sources, we also wanted to identify the bacteria that shifted after clindamycin treatment, regardless of colony source. By analyzing all mice that had sequence data from fecal samples collected at baseline and after clindamycin treatment, we identified `r clindamycin_otus_sig` OTUs and `r clindamycin_families_sig` families that were altered after clindamycin treatment (Fig. 6 and Tables S10-11). Interestingly, when we compared the list of significant clindamycin impacted bacteria with the bacteria that consistently varied across groups over the initial 3 timepoints of our experiment, we found 3 OTUs (*Lachnospiraceae* (OTU 130), *Lactobacillus* (OTU 6), *Enterococcus* (OTU 23)) and 3 families (*Porphyromonadaceae*, *Enterococcaceae*, *Lachnospiraceae*) overlapped (Fig. 5, Fig. 6C-D). These findings demonstrate that clindamycin has a consistent impact on the fecal bacterial communities of mice from all colony sources and only a subset of the taxa also varied across colony sources.

@@

#Venn diagram calculations for OTUs----
venn_otus_source <- length(dayn1to1_and_interp_combined)
venn_otus_intersect <- length(intersect(dayn1to1_and_interp_combined, paired_and_interp_combined))
venn_otus_clind <-  length(paired_and_interp_combined)
#Venn diagram calculations for families
venn_families_source <- length(dayn1to1_and_interp_combined_f)
venn_families_intersect <- length(intersect(dayn1to1_and_interp_combined_f, paired_and_interp_combined_f))
venn_families_clind <-  length(paired_and_interp_combined_f)
```
**Source-specific and clindamycin impacted bacteria distinguish *C. difficile* colonization status in mice.** After identifying taxa that varied by colony source, changed after clindamycin treatment, or both, we next wanted to determine which taxa were influencing the variation in *C. difficile* colonization at day 7 (Fig. 1D, Fig. S1C). We trained L2-regularized logistic regression models with input bacterial community data from the baseline, post-clindamycin, and post-infection timepoints of the experiment to predict *C. difficile* colonization status on day 7 (Fig. S3A-B). All models were better at predicting *C. difficile* colonization status on day 7 than random chance (all *P* $\le$ `r format(max_p_vs_random, scientific = TRUE, digits = 2)`; Table S12). However, the models trained with OTU level data generally performed better than those trained with family level data with the expception of the models based on the post-infection (day 1) communities (Fig. S3C-D). Interestingly, the model based on the post-clindamycin (day 0) community OTU data performed significantly better than all other models with an area under the receiving operator characteristic curve (AUROC) of `r median_otu_day0_AUROC` (*P*~FDR~ $\le$ `r format(max_padj_otu_day0_comp, scientific = TRUE, digits = 2)` for pairwise comparisons; Table S13). Thus, we were able to use community bacterial relative abundance data alone to differentiate mice that had cleared *C. difficile* before day 7 from the mice still colonized with *C. difficile* at that timepoint. Interestingly, the model built with OTU relative abundance data post-clindamycin treatment had the best performance, suggesting how the bacterial community responds to clindamycin treatment has the greatest influence on subsequent *C. difficile* colonization dynamics.

Next, to examine the bacteria that were driving each model’s performance, we pulled out the top 20 taxa that had the highest absolute feature weights in each of the 6 models (Tables S14-15). First, we looked at OTUs from the model with the best performance that was based on the post-clindamycin treatment bacterial community data. While most of the 20 OTUs had low relative abundances on day 0, *Enterobacteriace*, *Bacteroides* and *Proteus* had high relative abundances in at least one source of mice and significantly varied across sources (Fig. 7A). Next, the top 20 taxa from each model were compared to the list of taxa that varied across colony source at the same timepoint (Fig. 5 and Tables S8-9) and the taxa that were altered by clindamycin treatment (Fig. 6 and Table S10-11). We found a subset of OTUs and families that were important to the model and overlapped with bacteria that varied by either source, clindamycin treatment, or both (Fig. S4, S5A-C). Combining the overall results for the 3 OTU models identified `r venn_otus_source` OTUs associated with source, `r venn_otus_clind` OTUs associated with clindamycin treatment, and `r venn_otus_intersect` OTUs associated with both (Fig. 7B). Combining the overall results for the 3 family models identified `r venn_families_source` families associated with source, `r venn_families_clind` families associated with clindamycin treatment and `r venn_families_intersect` families associated with both (Fig. S5D). Several OTUs (*`r list_key_otus`*) and families (*`r list_key_families`*) appeared across at least 2 models, so we examined how the relative abundances of these key taxa varied over the course of the experiment (Fig. 8 and Fig. S6). Throughout the experiment, there was at least 1 timepoint where relative abundances of these taxa significantly varied across sources (Table S16-17). Interestingly, there were no taxa that emerged as consistently enriched or depleted in mice that were colonized past 7 days post-infection with *C. difficile* 630, suggesting multiple bacteria influence the time needed to clear the infection. Together, these results suggest the initial bacterial communities and their responses to clindamycin influence on the time needed to clear *C. difficile*.
